---
title: "Three models that need more than the sketch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three models that need more than the sketch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tidyABM)
```

Three of the models in `vignette("models")` run correctly and still do not show
the behaviour they are famous for. That is not a bug in the package or in the
translation. It is that the short version of each model, the one that fits in a
paragraph, leaves out the mechanism that produces the result. This vignette shows
what is missing in each case and puts it back.

The distinction matters when you are porting models. A translation can be
faithful to the source you were given and still be scientifically wrong, because
the source was a sketch. The only way to catch it is to check the output against
what the model is supposed to do, which is why every model in this package's test
suite asserts a behavioural claim rather than just "it ran".

## El Farol: one predictor is no predictor

Arthur's bar problem is famous for never settling. A hundred people decide
independently whether to go to a bar that is only fun below 60. If everyone
expects it to be empty everyone goes, and the expectation destroys itself.

The short version gives every agent the same forecast, last week's attendance,
and differs only in the threshold each will tolerate:

```{r}
short_pop <- abm_setup(
  agents  = abm_agents(n = 100, threshold = ~runif(n, 40, 80)),
  globals = list(last_attendance = 60))

short_go <- abm_go(
  abm_rules(go_today ~ last_attendance < threshold),
  abm_global(last_attendance ~ sum(go_today)))

short <- abm_run(short_pop, short_go, ticks = 40, seed = 2)

tail(abm_globals(short)$last_attendance, 12)
```

Everybody goes, then nobody goes, forever. With a single shared forecast the
population is effectively one agent, and one agent chasing its own tail produces a
two-cycle. Giving the agents heterogeneous *fixed* forecasts is not enough
either. It moves the cycle, but the map is still deterministic and still
converges.

Arthur's actual mechanism is **inductive**: each agent holds several candidate
predictors, scores them against what actually happened, and acts on whichever has
been working. Nobody's forecast is stable, so the population's response to a given
history keeps changing, and attendance never settles.

That is expressible here, but it needs each agent to carry a *set* of strategies,
and the grammar has no shorthand for that. One column per weight per strategy is
the only way. Ordinary R metaprogramming fills the gap:

```{r}
MEMORY <- 5    # weeks of attendance each predictor looks at
N_STRAT <- 10  # candidate predictors per agent
CAPACITY <- 60

# each strategy s is a linear rule: w_s0 * 100 + w_s1 * att1 + ... + w_s5 * att5
strategy_columns <- function() {
  cols <- list()
  for (s in seq_len(N_STRAT)) {
    for (j in 0:MEMORY) cols[[sprintf("w%d_%d", s, j)]] <- ~runif(n, -1, 1)
    cols[[sprintf("e%d", s)]] <- 0     # rolling error of strategy s
  }
  cols$active <- 1L                    # which one the agent currently trusts
  cols
}

farol <- abm_setup(
  agents  = do.call(abm_agents, c(list(n = 100), strategy_columns())),
  globals = setNames(as.list(rep(CAPACITY, MEMORY)), paste0("att", 1:MEMORY))
)
```

The steps themselves are ordinary: forecast, act, observe, re-score, switch.

```{r}
f <- function(lhs, rhs) rlang::new_formula(str2lang(lhs), str2lang(rhs))
lags <- paste0("att", 1:MEMORY)

forecast <- do.call(abm_rules, lapply(seq_len(N_STRAT), function(s) {
  f(paste0("p", s),
    paste(c(sprintf("w%d_0 * 100", s), sprintf("w%d_%d * %s", s, 1:MEMORY, lags)),
          collapse = " + "))
}))

act <- abm_rules(f("go_today", paste0(
  "case_when(", paste(sprintf("active == %d ~ p%d", 1:(N_STRAT - 1), 1:(N_STRAT - 1)),
                      collapse = ", "),
  ", TRUE ~ p", N_STRAT, ") < ", CAPACITY)))

observe <- do.call(abm_global, c(
  lapply(MEMORY:2, function(j) f(paste0("att", j), paste0("att", j - 1))),
  list(f("att1", "sum(go_today)"))
))

rescore <- do.call(abm_rules, lapply(seq_len(N_STRAT), function(s)
  f(paste0("e", s), sprintf("0.8 * e%d + 0.2 * abs(p%d - att1)", s, s))))

switch_to_best <- abm_rules(f("active", paste0(
  "case_when(",
  paste(sprintf("e%d <= pmin(%s) ~ %dL", 1:(N_STRAT - 1),
                vapply(1:(N_STRAT - 1),
                       function(s) paste0("e", setdiff(1:N_STRAT, s), collapse = ", "),
                       character(1)),
                1:(N_STRAT - 1)), collapse = ", "),
  ", TRUE ~ ", N_STRAT, "L)")))

go <- abm_go(forecast, act, observe, rescore, switch_to_best)

r <- abm_run(farol, go, ticks = 300, seed = 2)

attendance <- abm_globals(r)$att1[-(1:100)]
c(mean = round(mean(attendance), 1), sd = round(sd(attendance), 1),
  min = min(attendance), max = max(attendance),
  distinct = length(unique(attendance)))
```

Attendance now hovers near the capacity of 60 without ever repeating itself,
which is Arthur's result. Note how sensitive it is. With three predictors instead
of ten it locks up again. The fluctuation is a property of the predictor pool, not something
the model produces on its own.

**What this exposes about the grammar.** There is no compact way to give an agent
a *set* of anything. Ten strategies over five lags is seventy columns, and only
metaprogramming makes that bearable. A future `abm_agents()` might accept
matrix-valued or list-valued columns. Until then, build the spec with `do.call()`.

## Ethnocentrism: cooperation has to be conditional, and neighbours have to be kin

Hammond and Axelrod's model is famous for showing that in-group favouritism can
evolve from nothing. The short version gives agents a tag and a strategy, but the
strategy ignores the tag, so the tag does no work, defection is simply the best
move, and the population runs itself down.

The real model gives every agent **two** strategy bits: whether to cooperate with
someone of your own tag, and whether to cooperate with someone of a different one.
That makes four types.

```{r}
COST <- 0.01; BENEFIT <- 0.03; BASE_PTR <- 0.12; DEATH <- 0.10; CAPACITY <- 800

random_traits <- list(
  tag      ~ sample(c("red", "blue"), n(), replace = TRUE),
  coop_in  ~ sample(c(TRUE, FALSE),   n(), replace = TRUE),
  coop_out ~ sample(c(TRUE, FALSE),   n(), replace = TRUE)
)

start_pop <- function(network = NULL) {
  abm_setup(
    agents = abm_agents(
      n = 400,
      tag      = ~sample(c("red", "blue"), n, replace = TRUE),
      coop_in  = ~sample(c(TRUE, FALSE), n, replace = TRUE),
      coop_out = ~sample(c(TRUE, FALSE), n, replace = TRUE),
      ptr      = BASE_PTR),
    network = network)
}

label <- function(r) {
  dplyr::mutate(r, type = case_when(
    coop_in & !coop_out ~ "ethnocentric",
    coop_in &  coop_out ~ "altruist",
   !coop_in & !coop_out ~ "egoist",
    TRUE                ~ "traitor"))
}
shares <- function(r, t) {
  x <- label(r); x <- x$type[x$tick == t]
  round(sort(table(x) / length(x), decreasing = TRUE), 3)
}
```

Even with conditional strategies, a well-mixed population does not produce
ethnocentrism. Everyone meets a random stranger, so the tag carries no
information, and the cheapest strategy wins:

```{r}
well_mixed_go <- abm_go(
  abm_match(pair = "random"),
  abm_rules(give ~ if_else(partner_tag == tag, coop_in, coop_out)),
  abm_rules(ptr ~ BASE_PTR - COST * give + BENEFIT * partner_give),
  abm_birth(when = runif(n()) < ptr),
  abm_death(when = runif(n()) < DEATH + 0.25 * pmax(0, (n() - CAPACITY) / CAPACITY)),
  do.call(abm_birth, list(n = 8, cost = random_traits))
)

mixed <- abm_run(start_pop(), well_mixed_go, ticks = 400, seed = 1)

shares(mixed, 400)
```

Egoists, with ethnocentrics behind them. This is not a failure. It is Hammond
and Axelrod's own control condition, and it is the point of their paper:
ethnocentrism needs *local* structure. Offspring have to settle next to their parents, so that your
neighbours are disproportionately your kin, so that "same tag" actually predicts
"will cooperate with me".

That is what the model needs and it is the one thing the package could not
express: `attach_via` could put a newborn next to a random agent or a
well-connected one, but not next to its own parent. `from = "parent"` closes
that gap.

```{r}
local_go <- abm_go(
  abm_match(pair = "network"),
  abm_rules(give ~ if_else(partner_tag == tag, coop_in, coop_out)),
  abm_rules(ptr ~ BASE_PTR - COST * give + BENEFIT * partner_give),
  abm_birth(when = runif(n()) < ptr,
            attach_via = abm_match(pair = "network", from = "parent")),
  abm_death(when = runif(n()) < DEATH + 0.25 * pmax(0, (n() - CAPACITY) / CAPACITY)),
  do.call(abm_birth, list(n = 8, attach_via = abm_match(pair = "network"),
                          cost = random_traits))
)

local <- abm_run(start_pop(abm_network(type = "random", degree = 4)), local_go,
                 ticks = 400, seed = 1)

shares(local, 400)
```

Ethnocentrics now lead, and egoists are down to about a third of their
well-mixed share. The clustering is
visible directly in the network:

```{r}
edges <- abm_edges(local)
final <- local[local$tick == 400, ]
tags  <- setNames(final$tag, final$.id)
mean(tags[as.character(edges$from)] == tags[as.character(edges$to)], na.rm = TRUE)
```

Around 0.5 would mean neighbours are random with respect to tag. What you get is
close to 1: local reproduction has sorted the population into same-tag
neighbourhoods, and that is the whole mechanism.

## Zakah: nobody is poor in a model with no risk

The zakah model collects 2.5% from every household above the *nisab* and shares it
among households below a poverty line. As written it stops working after about ten
ticks:

```{r}
NISAB <- 100
consume <- abm_rules(wealth ~ wealth + income - (0.6 * income + 0.02 * wealth))

short_pop <- abm_setup(
  agents  = abm_agents(n = 500, wealth = ~rlnorm(n, 4, 0.5),
                       income = ~rlnorm(n, 3, 0.4)),
  globals = list(zakah_pool = 0))

short_go <- abm_go(
  consume,
  abm_global(zakah_pool ~ sum(if_else(wealth > NISAB, wealth * 0.025, 0))),
  abm_rules(wealth ~ if_else(wealth > NISAB, wealth * 0.975, wealth)),
  abm_rules(wealth ~ if_else(wealth < 30,
                             wealth + zakah_pool / sum(wealth < 30), wealth)))

short <- abm_run(short_pop, short_go, ticks = 100, seed = 12)

# how many households are below the poverty line, over time?
tapply(short$wealth < 30, short$tick, sum)[c("0", "5", "10", "50", "100")]
```

The pool keeps being collected and stops being distributed, so zakah becomes a
flat tax that goes nowhere.

Two things are wrong, and only one of them is the threshold.

**The poverty line is absolute in a model where wealth grows.** Everyone crosses
30 within a few ticks. Making it relative, at half the median, which is the
standard measure of relative poverty, is a one-line fix.

**More importantly, the model has no risk.** The consumption rule is
`wealth * 0.98 + 0.4 * income`, which converges to exactly `20 * income` for every
household. Wealth becomes a deterministic function of income, the distribution
gets *tighter* over time, and nobody is ever persistently poor no matter where you
put the line. A redistribution model with nothing to redistribute against is
measuring nothing.

Put in a source of dispersion, meaning income that moves plus occasional large
losses, and the model has something to say:

```{r}
shocks <- abm_rules(
  income ~ exp(0.9 * log(income) + 0.1 * 3 + rnorm(n(), 0, 0.15)),  # AR(1) in logs
  wealth ~ wealth - if_else(runif(n()) < 0.03, wealth * 0.6, 0)     # occasional hit
)
consume <- abm_rules(wealth ~ pmax(0.01, wealth + income - (0.6 * income + 0.02 * wealth)))
line    <- abm_global(poverty_line ~ 0.5 * median(wealth))

start <- function() abm_setup(
  agents  = abm_agents(n = 500, wealth = ~rlnorm(n, 4, 0.5), income = ~rlnorm(n, 3, 0.4)),
  globals = list(zakah_pool = 0, poverty_line = 30))

zakah_go <- abm_go(
  shocks, consume, line,
  abm_global(zakah_pool ~ sum(if_else(wealth > NISAB, wealth * 0.025, 0))),
  abm_rules(wealth ~ if_else(wealth > NISAB, wealth * 0.975, wealth)),
  abm_rules(wealth ~ if_else(wealth < poverty_line,
                             wealth + zakah_pool / pmax(1, sum(wealth < poverty_line)),
                             wealth))
)

baseline_go <- abm_go(shocks, consume, line)

with_zakah <- abm_run(start(), zakah_go,    ticks = 300, seed = 12)
baseline   <- abm_run(start(), baseline_go, ticks = 300, seed = 12)
```

```{r}
gini <- function(x) { x <- sort(x); n <- length(x); sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x)) }
compare <- function(r) {
  w <- r$wealth[r$tick == 300]
  c(p10 = round(quantile(w, 0.1), 1), median = round(median(w), 1),
    gini = round(gini(w), 3))
}
rbind(baseline = compare(baseline), zakah = compare(with_zakah))
```

The bottom decile is higher and the Gini coefficient is a third lower. The
recipient pool no longer empties:

```{r}
poor_share <- function(r) {
  pl <- abm_globals(r)$poverty_line[match(r$tick, abm_globals(r)$tick)]
  tapply(r$wealth < pl, r$tick, mean)
}
c(baseline = round(mean(tail(poor_share(baseline), 100)), 3),
  zakah    = round(mean(tail(poor_share(with_zakah), 100)), 3))
```

One caveat on reading that last number. Transferring 2.5% of nearly everyone's
wealth to the few per cent below the line is an enormous per-head transfer, so
relative poverty is not merely reduced but eliminated. Real zakah is levied on
*zakatable* assets held for a year and distributed across eight categories, of
which the poor are two. If you want the model to say something about actual policy,
that is the next thing to calibrate.

## What to take from this

The three failures have the same shape: a mechanism was compressed out of the
description, and the compressed version still runs.

* El Farol lost the *inductive* part, the agents that revise which forecast they
  trust. Without it the population is one agent.
* Ethnocentrism lost both the tag-conditional strategy and the local
  reproduction. Without the second one you reproduce the paper's control, not its
  result.
* Zakah lost the risk process. Without it there is no poverty to redistribute
  against, and the threshold question is a distraction from that.

Two of the three needed no change to the package. The third needed one argument,
`from = "parent"`, which was a genuine gap: there was no way to put an offspring
next to its parent, and that is a mechanism a lot of spatial and kin-structured
models depend on.
