---
title: "Three models that need more than the sketch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three models that need more than the sketch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tidyABM)
```

Three of the models in `vignette("models")` run correctly and still do not show
the behaviour they are famous for. That is not a bug in the package or in the
translation. It is that the short version of each model, the one that fits in a
paragraph, leaves out the mechanism that produces the result. This vignette shows
what is missing in each case and puts it back.

The distinction matters when you are porting models. A translation can be
faithful to the source you were given and still be scientifically wrong, because
the source was a sketch. The only way to catch it is to check the output against
what the model is supposed to do, which is why every model in this package's test
suite asserts a behavioural claim rather than just "it ran".

## El Farol: one predictor is no predictor

Arthur's bar problem is famous for never settling. A hundred people decide
independently whether to go to a bar that is only fun below 60. If everyone
expects it to be empty everyone goes, and the expectation destroys itself.

The short version gives every agent the same forecast, last week's attendance,
and differs only in the threshold each will tolerate:

```{r}
short_pop <- abm_setup(
  agents  = abm_agents(n = 100, threshold = ~runif(n, 40, 80)),
  globals = list(last_attendance = 60))

short_go <- abm_go(
  abm_rules(go_today ~ last_attendance < threshold),
  abm_global(last_attendance ~ sum(go_today)))

short <- abm_run(short_pop, short_go, ticks = 40, seed = 2)

tail(abm_globals(short)$last_attendance, 12)
```

Everybody goes, then nobody goes, forever. With a single shared forecast the
population is effectively one agent, and one agent chasing its own tail produces a
two-cycle. Giving the agents heterogeneous *fixed* forecasts is not enough
either. It moves the cycle, but the map is still deterministic and still
converges.

Arthur's actual mechanism is **inductive**: each agent holds several candidate
predictors, scores them against what actually happened, and acts on whichever has
been working. Nobody's forecast is stable, so the population's response to a given
history keeps changing, and attendance never settles.

Each agent therefore has to carry a *set* of predictors rather than a number,
and a column holds a set as readily as it holds a scalar. `w` is a matrix of
weights per agent, ten predictors over five lags and a constant; `e` is that
agent's rolling error for each of the ten; `p` is what each of them said this
week. All three are ordinary list columns.

```{r}
MEMORY <- 5    # weeks of attendance each predictor looks at
N_STRAT <- 10  # candidate predictors per agent
CAPACITY <- 60

farol <- abm_setup(
  agents = abm_agents(
    n = 100,
    w = ~lapply(seq_len(n), function(i)
          matrix(runif(N_STRAT * (MEMORY + 1), -1, 1), N_STRAT, MEMORY + 1)),
    e = ~lapply(seq_len(n), function(i) numeric(N_STRAT)),
    p = ~lapply(seq_len(n), function(i) numeric(N_STRAT)),
    go_today = FALSE),
  globals = as.list(setNames(rep(CAPACITY, MEMORY), paste0("att", 1:MEMORY))),
  seed = 1)
```

The tick is the four things the description names: forecast with every
predictor, act on one, observe what happened, re-score them all. There is no
fifth step for switching, because "act on whichever has been working" is
`which.min(e)` read at the moment of acting.

```{r}
go <- abm_go(
  abm_rules(p ~ lapply(w, function(W) as.vector(W %*% c(100, att1, att2, att3, att4, att5)))),
  abm_rules(go_today ~ mapply(function(pi, ei) pi[which.min(ei)], p, e) < CAPACITY),
  abm_global(att5 ~ att4, att4 ~ att3, att3 ~ att2, att2 ~ att1,
             att1 ~ sum(go_today)),
  abm_rules(e ~ mapply(function(ei, pi) 0.8 * ei + 0.2 * abs(pi - att1),
                       e, p, SIMPLIFY = FALSE))
)

r <- abm_run(farol, go, ticks = 300, seed = 2)

attendance <- abm_globals(r)$att1[-(1:101)]
c(mean = round(mean(attendance), 1), sd = round(sd(attendance), 1),
  min = min(attendance), max = max(attendance),
  distinct = length(unique(attendance)))
```

Attendance sits at the capacity of 60 and keeps moving, which is Arthur's
result. It is worth being precise about what "keeps moving" means, because the
short version did not merely fluctuate less, it locked into a cycle. Over the
last 200 ticks this run visits 16 distinct attendance levels and matches no lag
up to six, so there is no period to find.

The mechanism is the size of the predictor pool, not the inductive machinery on
its own. Across four population draws, ten predictors per agent never settles
(sd 1.8 to 3.7, ten to seventeen distinct levels, no period); three predictors
locks into a cycle in two draws of four; one predictor locks in all four, which
is the degenerate case the first run showed. Both seeds matter here:
`abm_setup(seed =)` fixes which predictors the agents are born with, and it
changes the answer as much as `abm_run(seed =)` does.

```{r}
one_predictor <- abm_setup(
  agents = abm_agents(
    n = 100,
    w = ~lapply(seq_len(n), function(i) matrix(runif(MEMORY + 1, -1, 1), 1, MEMORY + 1)),
    e = ~lapply(seq_len(n), function(i) numeric(1)),
    p = ~lapply(seq_len(n), function(i) numeric(1)),
    go_today = FALSE),
  globals = as.list(setNames(rep(CAPACITY, MEMORY), paste0("att", 1:MEMORY))),
  seed = 1)

tail(abm_globals(abm_run(one_predictor, go, ticks = 300, seed = 2))$att1, 8)
```

**What this exposes about the grammar.** Nothing, as it turns out, and that is
worth recording. This model was first written as seventy scalar columns and five
blocks of `rlang::new_formula()` scaffolding, and it was described here as the
model that proved the grammar could not give an agent a set. It can. What was
missing was the idiom, not the capability: rules over list columns are `lapply()`
and `mapply()` where you would otherwise write arithmetic, and nothing pointed
that out. The minority game and the naming game found it first.

## Ethnocentrism: cooperation has to be conditional, and neighbours have to be kin

Hammond and Axelrod's model is famous for showing that in-group favouritism can
evolve from nothing. The short version gives agents a tag and a strategy, but the
strategy ignores the tag, so the tag does no work, defection is simply the best
move, and the population runs itself down.

The real model gives every agent **two** strategy bits: whether to cooperate with
someone of your own tag, and whether to cooperate with someone of a different one.
That makes four types.

```{r}
COST <- 0.01; BENEFIT <- 0.03; BASE_PTR <- 0.12; DEATH <- 0.10; CAPACITY <- 800

random_traits <- list(
  tag      ~ sample(c("red", "blue"), n(), replace = TRUE),
  coop_in  ~ sample(c(TRUE, FALSE),   n(), replace = TRUE),
  coop_out ~ sample(c(TRUE, FALSE),   n(), replace = TRUE)
)

start_pop <- function(network = NULL) {
  abm_setup(
    agents = abm_agents(
      n = 400,
      tag      = ~sample(c("red", "blue"), n, replace = TRUE),
      coop_in  = ~sample(c(TRUE, FALSE), n, replace = TRUE),
      coop_out = ~sample(c(TRUE, FALSE), n, replace = TRUE),
      ptr      = BASE_PTR),
    network = network,
    seed    = 1)
}

label <- function(r) {
  dplyr::mutate(r, type = case_when(
    coop_in & !coop_out ~ "ethnocentric",
    coop_in &  coop_out ~ "altruist",
   !coop_in & !coop_out ~ "egoist",
    TRUE                ~ "traitor"))
}
shares <- function(r, t) {
  x <- label(r); x <- x$type[x$tick == t]
  round(sort(table(x) / length(x), decreasing = TRUE), 3)
}
```

Even with conditional strategies, a well-mixed population does not produce
ethnocentrism. Everyone meets a random stranger, so the tag carries no
information, and the cheapest strategy wins:

```{r}
well_mixed_go <- abm_go(
  abm_match(pair = "random"),
  abm_rules(give ~ if_else(partner_tag == tag, coop_in, coop_out)),
  abm_rules(ptr ~ BASE_PTR - COST * give + BENEFIT * partner_give),
  abm_birth(when = runif(n()) < ptr),
  abm_death(when = runif(n()) < DEATH + 0.25 * pmax(0, (n() - CAPACITY) / CAPACITY)),
  abm_birth(n = 8, inherit = random_traits)
)

mixed <- abm_run(start_pop(), well_mixed_go, ticks = 400, seed = 1)

shares(mixed, 400)
```

Egoists, at half the population. This is not a failure. It is Hammond and
Axelrod's own control condition, and it is the point of their paper:
ethnocentrism needs *local* structure. Offspring have to settle next to their
parents, so that your neighbours are disproportionately your kin, so that "same
tag" actually predicts "will cooperate with me".

That is what the model needs, and it took two arguments the package did not
have. `attach_via` could put a newborn next to a random agent or a
well-connected one, but not next to its own parent, which `from = "parent"`
fixed. And a newborn got exactly one edge, which is not a place in a
neighbourhood but a leaf hanging off one; `links` gives it the degree the model
means. Both births below take four links, the degree the population starts
with.

```{r}
local_go <- abm_go(
  abm_match(pair = "network"),
  abm_rules(give ~ if_else(partner_tag == tag, coop_in, coop_out)),
  abm_rules(ptr ~ BASE_PTR - COST * give + BENEFIT * partner_give),
  abm_birth(when = runif(n()) < ptr, links = 4,
            attach_via = abm_match(pair = "network", from = "parent")),
  abm_death(when = runif(n()) < DEATH + 0.25 * pmax(0, (n() - CAPACITY) / CAPACITY)),
  abm_birth(n = 8, links = 4, inherit = random_traits,
            attach_via = abm_match(pair = "network"))
)

local <- abm_run(start_pop(abm_network(type = "random", degree = 4)), local_go,
                 ticks = 400, seed = 1)

shares(local, 400)
```

Egoists collapse, from about half the population to about an eighth, and the two
strategies that cooperate with their own kind take three quarters of it between
them. That is the paper's result and it replicates: over five population draws
egoists average 0.37 well mixed and 0.15 locally.

Which of the two in-group strategies leads is a weaker claim than the vignette
used to make. Ethnocentrics average 0.37 against altruists' 0.33 across those
five draws, but they lead in only two of them, and this seed is one where the
altruists are marginally ahead. The reason is visible in the network: clustering
this strong means out-group encounters are rare, and a strategy that pays a cost
to refuse them saves little by doing so.

```{r}
edges <- abm_edges(local)
final <- local[local$tick == 400, ]
tags  <- setNames(final$tag, final$.id)
mean(tags[as.character(edges$from)] == tags[as.character(edges$to)], na.rm = TRUE)
```

Around 0.5 would mean neighbours are random with respect to tag. What you get is
around 0.84: local reproduction has sorted the population into same-tag
neighbourhoods, and that is the whole mechanism.

### What one edge per newborn was doing

`links` is not decoration, and the reason is worth showing, because the model
runs perfectly well without it and reports a *better* number.

With one edge per newborn, deaths prune four edges and births replace one. The
4-regular graph the run starts on is gone within twenty-five ticks. Over five
draws it ends at a mean degree of 0.97, with 27% of the population joined to
nobody at all and sitting out every match. What is left is a forest of
parent-child pairs, and 0.955 of those edges join agents of the same tag, which
is close to 1 for the trivial reason that most of them are a parent and its own
child.

With `links = 4` the network is a network: mean degree 3.3, 5% isolated, and a
same-tag share of 0.85 that describes neighbourhoods rather than kinship dyads.
The strategy shares barely move, 0.39 ethnocentrics against 0.37, which is the
good outcome: the mechanism was real, and the eroded network was overstating the
evidence for it.

## Zakah: nobody is poor in a model with no risk

The zakah model collects 2.5% from every household above the *nisab* and shares it
among households below a poverty line. As written it stops working after about ten
ticks:

```{r}
NISAB <- 100
consume <- abm_rules(wealth ~ wealth + income - (0.6 * income + 0.02 * wealth))

short_pop <- abm_setup(
  agents  = abm_agents(n = 500, wealth = ~rlnorm(n, 4, 0.5),
                       income = ~rlnorm(n, 3, 0.4)),
  globals = list(zakah_pool = 0))

short_go <- abm_go(
  consume,
  abm_global(zakah_pool ~ sum(if_else(wealth > NISAB, wealth * 0.025, 0))),
  abm_rules(wealth ~ if_else(wealth > NISAB, wealth * 0.975, wealth)),
  abm_rules(wealth ~ if_else(wealth < 30,
                             wealth + zakah_pool / sum(wealth < 30), wealth)))

short <- abm_run(short_pop, short_go, ticks = 100, seed = 12)

# how many households are below the poverty line, over time?
tapply(short$wealth < 30, short$tick, sum)[c("0", "5", "10", "50", "100")]
```

The pool keeps being collected and stops being distributed, so zakah becomes a
flat tax that goes nowhere.

Two things are wrong, and only one of them is the threshold.

**The poverty line is absolute in a model where wealth grows.** Everyone crosses
30 within a few ticks. Making it relative, at half the median, which is the
standard measure of relative poverty, is a one-line fix.

**More importantly, the model has no risk.** The consumption rule is
`wealth * 0.98 + 0.4 * income`, which converges to exactly `20 * income` for every
household. Wealth becomes a deterministic function of income, the distribution
gets *tighter* over time, and nobody is ever persistently poor no matter where you
put the line. A redistribution model with nothing to redistribute against is
measuring nothing.

Put in a source of dispersion, meaning income that moves plus occasional large
losses, and the model has something to say:

```{r}
shocks <- abm_rules(
  income ~ exp(0.9 * log(income) + 0.1 * 3 + rnorm(n(), 0, 0.15)),  # AR(1) in logs
  wealth ~ wealth - if_else(runif(n()) < 0.03, wealth * 0.6, 0)     # occasional hit
)
consume <- abm_rules(wealth ~ pmax(0.01, wealth + income - (0.6 * income + 0.02 * wealth)))
line    <- abm_global(poverty_line ~ 0.5 * median(wealth))

start <- function() abm_setup(
  agents  = abm_agents(n = 500, wealth = ~rlnorm(n, 4, 0.5), income = ~rlnorm(n, 3, 0.4)),
  globals = list(zakah_pool = 0, poverty_line = 30))

zakah_go <- abm_go(
  shocks, consume, line,
  abm_global(zakah_pool ~ sum(if_else(wealth > NISAB, wealth * 0.025, 0))),
  abm_rules(wealth ~ if_else(wealth > NISAB, wealth * 0.975, wealth)),
  abm_rules(wealth ~ if_else(wealth < poverty_line,
                             wealth + zakah_pool / pmax(1, sum(wealth < poverty_line)),
                             wealth))
)

baseline_go <- abm_go(shocks, consume, line)

with_zakah <- abm_run(start(), zakah_go,    ticks = 300, seed = 12)
baseline   <- abm_run(start(), baseline_go, ticks = 300, seed = 12)
```

```{r}
gini <- function(x) { x <- sort(x); n <- length(x); sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x)) }
compare <- function(r) {
  w <- r$wealth[r$tick == 300]
  c(p10 = round(quantile(w, 0.1), 1), median = round(median(w), 1),
    gini = round(gini(w), 3))
}
rbind(baseline = compare(baseline), zakah = compare(with_zakah))
```

The bottom decile is higher and the Gini coefficient is a third lower. The
recipient pool no longer empties:

```{r}
poor_share <- function(r) {
  pl <- abm_globals(r)$poverty_line[match(r$tick, abm_globals(r)$tick)]
  tapply(r$wealth < pl, r$tick, mean)
}
c(baseline = round(mean(tail(poor_share(baseline), 100)), 3),
  zakah    = round(mean(tail(poor_share(with_zakah), 100)), 3))
```

One caveat on reading that last number. Transferring 2.5% of nearly everyone's
wealth to the few per cent below the line is an enormous per-head transfer, so
relative poverty is not merely reduced but eliminated. Real zakah is levied on
*zakatable* assets held for a year and distributed across eight categories, of
which the poor are two. If you want the model to say something about actual policy,
that is the next thing to calibrate.

## What to take from this

The three failures have the same shape: a mechanism was compressed out of the
description, and the compressed version still runs.

* El Farol lost the *inductive* part, the agents that revise which forecast they
  trust. Without it the population is one agent.
* Ethnocentrism lost both the tag-conditional strategy and the local
  reproduction. Without the second one you reproduce the paper's control, not its
  result.
* Zakah lost the risk process. Without it there is no poverty to redistribute
  against, and the threshold question is a distraction from that.

Two of the three needed no change to the package. Ethnocentrism needed two
arguments, and both were the same missing idea: a newborn has to arrive
*somewhere*. `from = "parent"` says where, and `links` says how much of a
neighbourhood it gets when it lands. Without the first there is no kin
structure. Without the second the kin structure eats the network.

There is a fourth thing to take from this, which is about reading a result
rather than writing a model. Each of these corrections was checked by asking
whether the number the model reports means what the surrounding sentence claims.
The uncorrected El Farol had a plausible attendance series that was a two-cycle.
The uncorrected ethnocentrism had a same-tag edge share of 0.96 that was
measuring a forest of parent-child pairs. Both are the kind of number a model
happily produces and a reader happily accepts.
