---
title: "Model gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tidyABM)
```

These are the models tidyABM was designed against. Each one is here because it
demanded something the previous ones did not, so read them as a sequence: the
grammar grew to fit them, not the other way round.

## Simple economy

*Wilensky & Rand, ch. 2.* Money moves at random and inequality appears anyway.

```{r}
economy <- abm_setup(agents = abm_agents(n = 500, money = 100))

r <- abm_run(economy, abm_go(
  abm_match(pair = "random", role = list(giver = money > 0, receiver = TRUE)),
  abm_rules(money ~ if_else(.role == "giver", money - 1, money + 1))
), ticks = 500, seed = 1)

round(quantile(r$money[r$tick == 500]))
```

**What it introduced:** `role`. A transfer needs a direction, and the two
conditions in `role = list(giver = money > 0, receiver = TRUE)` say which agent
in each pair can take which side. A pair where neither can give is dropped.

## El Farol

*Arthur, 1994, simplified.* Everyone wants to go to the bar unless it is crowded,
and their decisions are what make it crowded.

```{r}
elfarol <- abm_setup(
  agents  = abm_agents(n = 100, threshold = ~runif(n, 40, 80)),
  globals = list(last_attendance = 60)
)

r <- abm_run(elfarol, abm_go(
  abm_rules(go_today ~ last_attendance < threshold),
  abm_global(last_attendance ~ sum(go_today))
), ticks = 20, seed = 2)

abm_globals(r)$last_attendance
```

**What it introduced:** `abm_global()`, and the fact that matching is optional —
this is the first model with no pairing at all.

**A warning about this version.** With one shared predictor, the model collapses
into a two-cycle: attendance overshoots, everyone stays home, attendance
undershoots, everyone goes. Arthur's result needs agents that hold several
candidate forecasts and switch to whichever has been working. The mechanism above
is the scaffolding, not the result — `vignette("corrections")` builds the working
version.

## Prisoner's dilemma, well mixed

Play, then copy whoever did better. Cooperation does not survive it.

```{r}
pd <- abm_setup(agents = abm_agents(
  n = 100,
  strategy = ~sample(c("cooperate", "defect"), n, replace = TRUE),
  payoff = 0
))

r <- abm_run(pd, abm_go(
  abm_match(pair = "random"),
  abm_rules(payoff ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ 3,
    strategy == "defect"    & partner_strategy == "defect"    ~ 1,
    strategy == "defect"    & partner_strategy == "cooperate" ~ 5,
    strategy == "cooperate" & partner_strategy == "defect"    ~ 0
  )),
  abm_match(pair = "random"),
  abm_rules(strategy ~ if_else(partner_payoff > payoff, partner_strategy, strategy))
), ticks = 100, seed = 3)

mean(r$strategy[r$tick == 100] == "cooperate")
```

**What it introduced:** multiple match phases in one tick. Playing and imitating
are two different pairings, written one after the other.

## Ethnocentrism

Cooperation with a cost of living, and agents that reproduce and die.

```{r}
ethno <- abm_setup(agents = abm_agents(
  n = 100,
  tag      = ~sample(c("red", "blue"), n, replace = TRUE),
  strategy = ~sample(c("cooperate", "defect"), n, replace = TRUE),
  resource = 10
))

r <- abm_run(ethno, abm_go(
  abm_match(pair = "random"),
  abm_rules(resource ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ resource + 2,
    strategy == "defect"    & partner_strategy == "cooperate" ~ resource + 4,
    strategy == "cooperate" & partner_strategy == "defect"    ~ resource - 1,
    TRUE ~ resource
  ) - 1),
  abm_birth(when = resource > 20, cost = resource ~ resource / 2),
  abm_death(when = resource <= 0)
), ticks = 30, seed = 4)

as.integer(table(r$tick))[c(1, 11, 21, 31)]
```

**What it introduced:** `abm_birth()` and `abm_death()`, as two more flat steps
in the sequence rather than a special population mechanism.

**A warning about this version.** The tag does no work: strategies are
unconditional, so cooperating is simply dominated and the population runs itself
down. Hammond and Axelrod's agents decide *separately* whether to cooperate with
their own tag and with others — and even that is not enough without local
reproduction. `vignette("corrections")` builds both variants and shows why the
second one matters.

## Rumour mill

A three-state diffusion where the rumour burns out before everyone hears it.

```{r}
rumour <- abm_setup(agents = abm_agents(
  n = 200, state = ~c("spreader", rep("ignorant", n - 1))
))

r <- abm_run(rumour, abm_go(
  abm_match(pair = "random"),
  abm_rules(state ~ case_when(
    state == "ignorant" & partner_state == "spreader" ~ "spreader",
    state == "spreader" & partner_state == "spreader" ~ "stifler",
    state == "spreader" & partner_state == "stifler"  ~ "stifler",
    TRUE ~ state
  ))
), ticks = 100, seed = 5)

table(r$state[r$tick == 100])
```

**What it introduced:** nothing new — which was the point. Pairwise state
comparison through `partner_*` columns is self-consistent by construction, so it
needs no `role`. Transfers of a conserved quantity need `role`; symmetric state
changes do not.

## Party

Segregation with no geography: agents move in opinion space until their nearest
neighbour is close enough.

```{r}
party <- abm_setup(agents = abm_agents(n = 200, opinion = ~runif(n, 0, 1)))

r <- abm_run(party, abm_go(
  abm_match(pair = "nearest", by = opinion),
  abm_rules(opinion ~ if_else(abs(opinion - partner_opinion) > 0.05,
                              runif(1), opinion))
), ticks = 100, seed = 6)

mean_gap <- function(x) mean(diff(sort(x)))
c(start = mean_gap(r$opinion[r$tick == 0]),
  end   = mean_gap(r$opinion[r$tick == 100]))
```

**What it introduced:** `pair = "nearest"`. Note that this pairing is
*directional* — your nearest neighbour need not have you as theirs — so each
agent gets its own `.group_id`, and `runif(1)` in the rule above is drawn once per
agent rather than once for the population.

## Market

*After Primer's "Simulating Supply and Demand".* Buyers and sellers with private
valuations haggle, and the price finds its own level.

```{r}
market <- abm_setup(agents = list(
  buyers  = abm_agents(n = 200, wtp = ~rnorm(n, 50, 10), offer = ~wtp * 0.8),
  sellers = abm_agents(n = 200, wta = ~rnorm(n, 40, 10), ask   = ~wta * 1.2)
))

r <- abm_run(market, abm_go(
  abm_match(pair = "opposite_group", by = .group, resolve = "negotiate",
            rounds = 5, positions = c(offer, ask), limits = c(wtp, wta)),
  abm_rules(offer ~ if_else(traded, offer * 0.98, offer * 1.02)),
  abm_rules(ask   ~ if_else(traded, ask * 1.02,   ask * 0.98))
), ticks = 200, seed = 7)

price <- tapply(r$price, r$tick, function(z) mean(z, na.rm = TRUE))
round(price[c("1", "50", "200")], 2)
```

**What it introduced:** several agent groups, and rule routing by column
existence. The `offer` rule mentions a column only buyers have, so it applies only
to buyers; the `ask` rule only to sellers. No `type ==` test, no merging of the two
tibbles.

## Voter model

Opinions spread along fixed connections rather than through the whole population.

```{r}
voter <- abm_setup(
  agents  = abm_agents(n = 100, opinion = ~sample(c(0, 1), n, replace = TRUE)),
  network = abm_network(type = "random", degree = 4)
)

r <- abm_run(voter, abm_go(
  abm_match(pair = "network"),
  abm_rules(opinion ~ partner_opinion)
), ticks = 200, seed = 8)

round(c(start = mean(r$opinion[r$tick == 0]),
        end   = mean(r$opinion[r$tick == 200])), 2)
```

**What it introduced:** `abm_network()` and `pair = "network"`. The network is a
separate edge list stored alongside the agent tibble, not a list-column on it.

## Public goods game

Groups of four, a pot that grows and is split evenly whoever paid in.

```{r}
pgg <- abm_setup(agents = abm_agents(
  n = 100, contribution = ~sample(c(0, 1), n, replace = TRUE), payoff = 0
))

r <- abm_run(pgg, abm_go(
  abm_match(pair = "random", size = 4),
  abm_rules(payoff ~ sum(contribution) * 2 / 4)
), ticks = 20, seed = 9)

table(r$payoff[r$tick == 20])
```

**What it introduced:** `size`, generalising pairs to groups. Rules reuse dplyr's
grouped-mutate semantics rather than adding group-aggregate functions of their
own, so `sum(contribution)` just means what you would expect.

## Iterated prisoner's dilemma

Same game, same partner every round, tit for tat. Cooperation survives.

```{r}
ipd <- abm_setup(
  agents  = abm_agents(n = 100, move = "cooperate", payoff = 0),
  network = abm_network(type = "random", degree = 1)
)

r <- abm_run(ipd, abm_go(
  abm_match(pair = "network"),
  abm_rules(payoff ~ case_when(
    move == "cooperate" & partner_move == "cooperate" ~ 3,
    move == "defect"    & partner_move == "defect"    ~ 1,
    move == "defect"    & partner_move == "cooperate" ~ 5,
    move == "cooperate" & partner_move == "defect"    ~ 0
  )),
  abm_rules(move ~ partner_move)
), ticks = 50, seed = 10)

table(r$move[r$tick == 50])
```

**What it introduced:** nothing. It is a composition of `abm_network(degree = 1)`
for fixed partnerships and the `partner_*` convention, which already carries a
partner's previous move forward. Memory of one round comes free; anything longer
needs an explicit column.

## Preferential attachment

*Wilensky & Rand, ch. 5.* The network grows, and hubs appear.

```{r}
pa <- abm_setup(
  agents  = abm_agents(n = 2),
  network = abm_network(type = "manual", edges = data.frame(from = 1, to = 2))
)

r <- abm_run(pa, abm_go(
  abm_birth(n = 1, attach_via = abm_match(pair = "network", from = "random_edge"))
), ticks = 300, seed = 11)

deg <- table(c(abm_edges(r)$from, abm_edges(r)$to))
c(max_degree = max(deg), median_degree = median(deg))
```

**What it introduced:** the only way the network changes during a run.
`from = "random_edge"` reuses NetLogo's trick — pick an edge uniformly, then one
of its endpoints — so selection is degree-proportional without anyone storing a
degree.

## Zakah redistribution

Consumption erodes wealth; those above the *nisab* pay 2.5% into a pool; those
below the poverty line share it.

```{r}
nisab <- 100
poverty_line <- 30

zakah <- abm_setup(
  agents  = abm_agents(n = 300, wealth = ~rlnorm(n, 4, 0.5),
                       income = ~rlnorm(n, 3, 0.4)),
  globals = list(zakah_pool = 0)
)

r <- abm_run(zakah, abm_go(
  abm_rules(wealth ~ wealth + income - (0.6 * income + 0.02 * wealth)),
  abm_global(zakah_pool ~ sum(if_else(wealth > nisab, wealth * 0.025, 0))),
  abm_rules(wealth ~ if_else(wealth > nisab, wealth * 0.975, wealth)),
  abm_rules(wealth ~ if_else(wealth < poverty_line,
                             wealth + zakah_pool / sum(wealth < poverty_line),
                             wealth))
), ticks = 50, seed = 12)

round(tail(abm_globals(r)$zakah_pool, 3), 1)
```

**What it introduced:** nothing mechanically — it is the first model made of
nothing but individual and population-level steps, which confirmed matching is
optional machinery rather than a requirement. Collection is modelled as a pool
rather than peer-to-peer transfers, both because that is how zakah institutions
actually work and because it sidesteps matching a mismatched number of payers and
recipients.

**Watch the two thresholds — and the missing risk.** `nisab` and `poverty_line`
are deliberately different, so a middle group pays nothing and receives nothing.
But the recipient pool empties within about ten ticks: the consumption rule is
mean-reverting, so every household converges on `20 * income` and nobody is
persistently poor. Zakah then becomes a pure drag on the payers. Check
`sum(wealth < poverty_line)` over the run before reading anything into the
inequality path; `vignette("corrections")` fixes it.

## Bank reserves

*Wilensky, NetLogo Sample Models.* A bank can only lend what it has, and whoever
asks first gets it.

```{r}
reserve_ratio <- 0.1

bankres <- abm_setup(
  agents  = abm_agents(n = 100, wallet = ~runif(n, 0, 50), savings = 0, loan = 0,
                       draw = 0),
  globals = list(bank_deposits = 0, bank_loans = 0, bank_reserves = 0)
)

r <- abm_run(bankres, abm_go(
  abm_match(pair = "random", role = list(giver = TRUE, receiver = TRUE)),
  abm_rules(gift ~ sample(c(0, 2, 5), 1)),
  abm_rules(wallet ~ if_else(.role == "giver", wallet - gift, wallet + partner_gift)),
  abm_rules(
    savings ~ if_else(wallet > 0, savings + wallet, savings - pmin(savings, abs(wallet))),
    wallet  ~ if_else(wallet > 0, 0, wallet + pmin(savings, abs(wallet)))
  ),
  abm_global(bank_deposits ~ sum(savings)),
  abm_sequential(
    draw          ~ if_else(wallet < 0 & bank_reserves > 0,
                            pmin(-wallet, bank_reserves), 0),
    loan          ~ loan + draw,
    wallet        ~ wallet + draw,
    bank_reserves ~ bank_reserves - draw
  ),
  abm_global(bank_loans    ~ sum(loan)),
  abm_global(bank_reserves ~ bank_deposits * reserve_ratio - bank_loans)
), ticks = 100, seed = 13)

round(tail(abm_globals(r), 3), 2)
```

**What it introduced:** `abm_sequential()`. Every earlier model works fine with
simultaneous updates, because a shared pool that is only ever *divided* — the
public goods pot, the zakah pool — gives the same answer either way. Lendable
reserves are different: they are *depleted*, so the first borrower has to change
what the second one sees. The step is scoped narrowly to that case — during the
loop a rule reads and writes its own agent's columns and any global, and nothing
else.

Note that the last `abm_sequential()` rule writes to a global. That is what makes
the depletion visible to the next agent; without it the reserve would only be
recomputed at the end of the tick and every agent would borrow against the same
balance.

The first rule is doing something too. Rules inside `abm_sequential()` cascade —
each one sees what the rule above it wrote — so the amount actually drawn is
worked out once, into `draw`, and the three rules that move money all refer to
it. Writing the condition out three times would not work: the second rule
brings `wallet` back to zero, and the third would then find nothing left to
test.
