---
title: "Getting started with tidyABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tidyABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tidyABM)
```

An agent-based model is usually written as nested loops: for each tick, for each
agent, do something. tidyABM asks for a *specification* instead. You declare what
the population is and list what happens to it, in order, and the package runs the
loops.

Three things come out of that. The model reads as a description of the mechanism
rather than as bookkeeping. Mistakes in the structure are caught when you build
the model, not on tick 400. And the output is tidy data, so analysing a run is
ordinary dplyr.

## The three parts

Every model in this package is the same three functions, written as three
statements:

1. **the world**, `abm_setup()`. The agents, the network, the globals.
2. **the tick**, `abm_go()`. The ordered steps replayed once per tick.
3. **the run**, `abm_run()`. The two above, plus `ticks`, `seed` and `record`.

```{r, eval = FALSE}
world  <- abm_setup(...)                                # 1
go     <- abm_go(...)                                   # 2
result <- abm_run(world, go, ticks = ..., seed = ...)   # 3
```

The first two return objects you can print and inspect, so every example below
builds them on their own lines rather than inside the call to `abm_run()`. That
is what lets you reuse one go block across several worlds, or run one world
through several go blocks.

Everything else sits inside one of the three: `abm_agents()` and
`abm_network()` are what you hand to `abm_setup()`, and every other `abm_*()`
function is a step inside `abm_go()`.

## The whole game

Wilensky and Rand's *Simple Economy* is the smallest interesting model there is.
Five hundred agents start with \$100 each. Every tick, everyone who has money
gives \$1 to someone else. Nothing else happens.

```{r}
# 1. the world
economy <- abm_setup(agents = abm_agents(n = 500, money = 100))
economy
```

`abm_setup()` declares the world. `abm_agents()` declares one group of agents
inside it. A plain value like `money = 100` is recycled across every agent.

Now the behaviour:

```{r}
# 2. the tick
go <- abm_go(
  abm_match(pair = "random", role = list(giver = money > 0, receiver = TRUE)),
  abm_rules(money ~ if_else(.role == "giver", money - 1, money + 1))
)
go
```

Two steps. `abm_match()` shuffles the population into pairs and works out who in
each pair is the giver. The `role` list says the giver must satisfy `money > 0`
and the receiver can be anyone, so a pair where neither agent has money is
dropped for this tick. Then `abm_rules()` moves the dollar.

```{r}
# 3. the run
result <- abm_run(economy, go, ticks = 500, seed = 1)
result
```

The result is one row per agent per tick. Tick 0 is the state `abm_setup()`
produced, before anything has run.

```{r}
start <- result$money[result$tick == 0]
end   <- result$money[result$tick == 500]

c(total_start = sum(start), total_end = sum(end))
round(summary(end))
```

Money is conserved, because no rule creates or destroys any, but the
distribution has gone from a spike at 100 to something with a long right tail.
That is the whole point of the model.

## Setup: agents, networks, globals

`abm_agents()` takes `n` and any number of named columns. A plain value recycles.
A one-sided formula is evaluated once and must return `n` values, or one value to
recycle. Formulas run in order, so a later column can use an earlier one:

```{r}
abm_agents(
  n = 200,
  wtp   = ~rnorm(n, 50, 10),
  offer = ~wtp * 0.8
)
```

A model can have several kinds of agent, declared as a named list. Each keeps its
own columns:

```{r}
market <- abm_setup(agents = list(
  buyers  = abm_agents(n = 5, wtp = ~rnorm(n, 50, 10), offer = ~wtp * 0.8),
  sellers = abm_agents(n = 5, wta = ~rnorm(n, 40, 10), ask   = ~wta * 1.2)
))
market
```

Agent ids are unique across the whole model, and every agent carries a `.group`
column naming its group.

Values shared by the whole population go in `globals`: a bar's last attendance,
a bank's ledger. They are readable inside every rule:

```{r}
elfarol <- abm_setup(
  agents  = abm_agents(n = 100, threshold = ~runif(n, 40, 80)),
  globals = list(last_attendance = 60)
)
```

A network is an edge list built once and kept for the whole run:

```{r}
abm_network(type = "random", degree = 4)
```

`type = "random"` gives every agent *exactly* `degree` neighbours, using
`igraph::sample_k_regular()`. `degree = 1` therefore pairs the whole population
off permanently.

## Go: a sequence of steps

`abm_go()` takes steps in order and dispatches them by type, not by argument
name. That means a model with several phases per tick is written flat:

```{r}
pd_go <- abm_go(
  # phase 1: play
  abm_match(pair = "random"),
  abm_rules(payoff ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ 3,
    strategy == "defect"    & partner_strategy == "defect"    ~ 1,
    strategy == "defect"    & partner_strategy == "cooperate" ~ 5,
    strategy == "cooperate" & partner_strategy == "defect"    ~ 0
  )),
  # phase 2: imitate whoever did better
  abm_match(pair = "random"),
  abm_rules(strategy ~ if_else(partner_payoff > payoff, partner_strategy, strategy))
)
pd_go
```

The sequence is checked when you build it. Two matches in a row, or a sequence
ending on a match, means a pairing nobody uses, and that is an error rather than a
silent waste:

```{r, error = TRUE}
abm_go(
  abm_match(pair = "random"),
  abm_match(pair = "random"),
  abm_rules(x ~ 1)
)
```

Matching is optional. A model where every agent decides alone needs no
`abm_match()` at all:

```{r}
farol_go <- abm_go(
  abm_rules(go_today ~ last_attendance < threshold),
  abm_global(last_attendance ~ sum(go_today))
)
abm_run(elfarol, farol_go, ticks = 5, seed = 1) |> abm_globals()
```

## What a match gives a rule

For a pair, every rule downstream of the match can see `partner_<col>` for each
of the partner's columns:

```{r}
rumour <- abm_setup(
  agents = abm_agents(n = 100, state = ~c("spreader", rep("ignorant", n - 1)))
)

rumour_go <- abm_go(
  abm_match(pair = "random"),
  abm_rules(state ~ case_when(
    state == "ignorant" & partner_state == "spreader" ~ "spreader",
    state == "spreader" & partner_state == "spreader" ~ "stifler",
    state == "spreader" & partner_state == "stifler"  ~ "stifler",
    TRUE ~ state
  ))
)

r <- abm_run(rumour, rumour_go, ticks = 100, seed = 1)

table(r$state[r$tick == 100])
```

For a group, rules are evaluated group by group, so an aggregate inside a rule
means "across this agent's group". The public goods game is one line because of
it:

```{r}
pgg <- abm_setup(agents = abm_agents(
  n = 100, contribution = ~sample(c(0, 1), n, replace = TRUE), payoff = 0
))

pgg_go <- abm_go(
  abm_match(pair = "random", size = 4),
  abm_rules(payoff ~ sum(contribution) * 2 / 4)
)

r <- abm_run(pgg, pgg_go, ticks = 10, seed = 1)

table(r$payoff[r$tick == 10])
```

Grouping is also what makes randomness inside a rule behave: with a match
standing, `sample(x, 1)` is drawn once per pair, not once for everybody.

## Simultaneous or sequential

Every rule in one `abm_rules()` call sees the state at the *start* of the step.
That is the synchronous update agent-based models normally assume, and it is the
one place `abm_rules()` deliberately differs from `dplyr::mutate()`:

```{r}
swap <- abm_setup(agents = abm_agents(n = 3, a = 1, b = 2))
swap_go <- abm_go(abm_rules(a ~ b, b ~ a))

abm_run(swap, swap_go, ticks = 1)
```

Sometimes that is wrong. If agents compete for a pool that gets *used up*, the
first agent to act has to change what the next one sees. `abm_sequential()`
processes agents one at a time in shuffled order, and its writes to globals are
visible to the agents after it:

```{r}
pot <- abm_setup(agents = abm_agents(n = 20, got = 0), globals = list(pot = 10))

pot_go <- abm_go(abm_sequential(
  got ~ if_else(pot > 0, 1, 0),
  pot ~ if_else(pot > 0, pot - 1, pot)
))

seq_run <- abm_run(pot, pot_go, ticks = 1, seed = 1)

sum(seq_run$got[seq_run$tick == 1])  # exactly the 10 units that existed
```

With `abm_rules()` instead, all twenty agents would see a full pot and all twenty
would be served. Use `abm_sequential()` only when that difference is the point:
it is slower and harder to reason about.

The order is a fresh shuffle unless you name one. When the order is itself part
of the model, say a queue at a counter or a sequential-service constraint, pass
`.order =` and the agents go through in its ascending order instead:

```{r}
queue <- abm_setup(agents = abm_agents(n = 20, place = ~sample(n), got = 0),
                   globals = list(pot = 10))

queue_go <- abm_go(abm_sequential(
  got ~ if_else(pot > 0, 1, 0),
  pot ~ if_else(pot > 0, pot - 1, pot),
  .order = place
))

q <- abm_run(queue, queue_go, ticks = 1, seed = 1)
sum(q$got[q$tick == 1 & q$place <= 10])   # the front of the queue, every time
```

## Repeating a block

A tick is one pass through `abm_go()`. Some models have a *phase* inside the
tick that has to finish first: an epidemic that burns out before anyone
reconsiders vaccinating, a round of proposals that runs until nobody is
rejected. `abm_repeat()` replays a block until a condition holds:

```{r}
grow <- abm_setup(agents = abm_agents(n = 5, x = 0))
grow_go <- abm_go(
  abm_repeat(abm_rules(x ~ x + 1), until = mean(x) >= 4, max = 100)
)

r <- abm_run(grow, grow_go, ticks = 1, seed = 1)
unique(r$x[r$tick == 1])
```

`until` is checked after each pass, so the block always runs at least once, and
`max` is required, because a condition that never becomes true would hang the
run.

## Neighbourhoods

A match gives an agent one partner. `abm_neighbours()` gives it an aggregate over
the agents *around* it, which is what a model needs when the question is "how
many of them" rather than "what did mine do". Each rule is
`column ~ aggregate`, evaluated over the neighbours' rows, with the focal
agent's own columns visible as `own_<col>` so that comparisons are expressible:

```{r}
town <- abm_setup(
  agents  = abm_agents(n = 40, wealth = ~round(runif(n, 0, 100))),
  network = abm_network(type = "poisson", degree = 4),
  seed    = 1
)
town_go <- abm_go(
  abm_neighbours(richer ~ sum(wealth > own_wealth))
)

r <- abm_run(town, town_go, ticks = 1, seed = 1)
table(r$richer[r$tick == 1], useNA = "ifany")
```

The neighbourhood is the network by default. `within =` makes it a neighbourhood
in *attribute space* instead, meaning everybody whose columns satisfy a condition,
network or no network. Hegselmann and Krause's bounded confidence model is one
step:

```{r}
crowd <- abm_setup(agents = abm_agents(n = 200, opinion = ~runif(n)),
                   globals = list(eps = 0.15), seed = 42)
crowd_go <- abm_go(
  abm_neighbours(opinion ~ mean(opinion),
                 within = abs(opinion - own_opinion) <= eps)
)

hk <- abm_run(crowd, crowd_go, ticks = 30)
round(sort(unique(round(hk$opinion[hk$tick == 30], 4))), 3)
```

One difference is worth knowing: an agent is inside its own *attribute*
neighbourhood whenever the condition holds of it, since "the mean opinion of
everyone I take seriously" includes its own, and never inside its own *network*
one.
Write `within = ... & .id != own_.id` if you want it out.

Two `abm_neighbours()` passes draw their random numbers independently, which is
right when each is a separate event and wrong when it is one event seen from two
sides. `abm_draw()` puts the draw on the edge instead, where both endpoints read
the same number:

```{r}
pairs <- abm_setup(agents = abm_agents(n = 30, met = 0L),
                   network = abm_network(type = "poisson", degree = 4),
                   seed = 2)
pairs_go <- abm_go(
  abm_draw(happened ~ runif(n()) < 0.5),
  abm_neighbours(met ~ sum(happened))
)

r <- abm_run(pairs, pairs_go, ticks = 1, seed = 1)
# every meeting is counted by both of the agents in it, so the total is even
sum(r$met[r$tick == 1], na.rm = TRUE)
```

`.each = "endpoint"` gives the two ends one draw each instead, read as `name`
from an agent's own side and `name_back` from the other, which is what an
asymmetric interaction needs, since whether I noticed you and whether you noticed
me are different questions about the same edge.

## Changing the population

`abm_birth()` and `abm_death()` are the only steps that add or remove agents.

```{r}
ethno <- abm_setup(agents = abm_agents(
  n = 100,
  strategy = ~sample(c("cooperate", "defect"), n, replace = TRUE),
  resource = 10
))

ethno_go <- abm_go(
  abm_match(pair = "random"),
  abm_rules(resource ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ resource + 2,
    strategy == "defect"    & partner_strategy == "cooperate" ~ resource + 4,
    strategy == "cooperate" & partner_strategy == "defect"    ~ resource - 1,
    TRUE ~ resource
  ) - 1),
  abm_birth(when = resource > 20, cost = resource ~ resource / 2),
  abm_death(when = resource <= 0)
)

r <- abm_run(ethno, ethno_go, ticks = 20, seed = 1)

table(r$tick)[c(1, 11, 21)]
```

`cost` says what reproduction costs, as ordinary `column ~ expression` formulas
applied to the parent and the newborn alike, so halving a resource splits it
between them.

One parent, one offspring, unless `times` says otherwise. It takes an expression
evaluated in the parent's row, so a fertility can be a column, a number, or a
draw:

```{r}
seeds <- abm_setup(agents = abm_agents(n = 10, age = 5), seed = 1)
seeds_go <- abm_go(
  abm_birth(when = age > 0, times = rpois(n(), 2), inherit = age ~ 0)
)

r <- abm_run(seeds, seeds_go, ticks = 1, seed = 1)
table(r$tick)
```

Each offspring is a row of its own before `inherit` is evaluated, so a mutation
drawn there differs from sibling to sibling rather than being drawn once and
copied.

## Reproducibility

Agent-based models are stochastic, so `seed` is an argument to `abm_run()` rather
than something you arrange yourself. It is set locally, so your session's random
state is untouched:

```{r}
set.seed(1); before <- runif(1)
invisible(abm_run(economy, go, ticks = 5, seed = 99))
set.seed(1); after <- runif(1)
identical(before, after)
```

Two runs of the same model with the same seed are identical:

```{r}
a <- abm_run(economy, go, ticks = 20, seed = 42)
b <- abm_run(economy, go, ticks = 20, seed = 42)
identical(as.data.frame(a), as.data.frame(b))
```

There is a sharp edge here worth knowing about. `abm_run()`'s seed fixes the
*run*, not the *model*. If your agents' starting columns are random, they were
drawn when `abm_setup()` was called, before `abm_run()` had a chance to set
anything, so two experiments built the same way can start from different
populations:

```{r}
random_pop <- function() abm_setup(agents = abm_agents(n = 50, x = ~runif(n)))
noop_go <- abm_go(abm_rules(x ~ x))

set.seed(1); one <- abm_run(random_pop(), noop_go, ticks = 1, seed = 5)
set.seed(2); two <- abm_run(random_pop(), noop_go, ticks = 1, seed = 5)
identical(one$x, two$x)
```

Seed both, and the whole experiment reproduces:

```{r}
experiment <- function() {
  m <- abm_setup(agents = abm_agents(n = 50, x = ~runif(n)), seed = 4)
  noop_go <- abm_go(abm_rules(x ~ x))
  abm_run(m, noop_go, ticks = 1, seed = 5)
}
set.seed(1); one <- experiment()
set.seed(2); two <- experiment()
identical(one$x, two$x)
```

One further caveat: a seed fixes a run given a version of the package, not across
versions. Reordering an internal random draw changes the exact numbers without
changing the model, so compare runs by their statistical behaviour rather than by
exact equality when you upgrade.

## How much to record

By default `abm_run()` keeps every agent of every tick, which is right for a
fixed population and wrong for a growing one. A run that ends with fifty
thousand agents has been keeping every one of them since the start. `record` says
how much to keep:

```{r}
sizes <- c(
  all      = nrow(abm_run(economy, go, ticks = 20, record = "all")),
  every_5  = nrow(abm_run(economy, go, ticks = 20, record = 5)),
  final    = nrow(abm_run(economy, go, ticks = 20, record = "final")),
  globals  = nrow(abm_run(economy, go, ticks = 20, record = "globals"))
)
sizes
```

A whole number keeps every *n*th tick plus the two ends. `"final"` keeps the
last tick only, and `"globals"` keeps none of the populations. Globals are
recorded every tick whatever you say, since they are one row each, so
`"globals"` is the setting for a model whose output is an aggregate. The run
itself is unchanged either way: the same seed gives the same final state at any
setting.

## Where to go next

`vignette("models")` works through the models the package was designed against,
each with the mechanism it demonstrates.
