---
title: "Getting started with tidyABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tidyABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(tidyABM)
```

An agent-based model is usually written as nested loops: for each tick, for each
agent, do something. tidyABM asks you to write the model as a *specification*
instead — a declaration of what the population is and an ordered list of what
happens to it — and runs the loops for you.

That buys three things. The model reads as a description of the mechanism rather
than as bookkeeping; mistakes in the structure are caught once, when you build the
model, rather than on tick 400; and the output is tidy data, so analysing a run is
ordinary dplyr.

## The whole game

Wilensky and Rand's *Simple Economy* is the smallest interesting model there is.
Five hundred agents start with \$100 each. Every tick, everyone who has money
gives \$1 to someone else. Nothing else happens.

```{r}
economy <- abm_setup(agents = abm_agents(n = 500, money = 100))
economy
```

`abm_setup()` declares the world; `abm_agents()` declares one group of agents
within it. A plain value like `money = 100` is recycled across every agent.

Now the behaviour:

```{r}
go <- abm_go(
  abm_match(pair = "random", role = list(giver = money > 0, receiver = TRUE)),
  abm_rules(money ~ if_else(.role == "giver", money - 1, money + 1))
)
go
```

Two steps. `abm_match()` shuffles the population into pairs and works out who in
each pair is the giver — the `role` list says the giver must satisfy
`money > 0` and the receiver can be anyone, so a pair where neither agent has
money is dropped for this tick. `abm_rules()` then moves the dollar.

```{r}
result <- abm_run(economy, go, ticks = 500, seed = 1)
result
```

The result is one row per agent per tick. Tick 0 is the state `abm_setup()`
produced, before anything has run.

```{r}
start <- result$money[result$tick == 0]
end   <- result$money[result$tick == 500]

c(total_start = sum(start), total_end = sum(end))
round(summary(end))
```

Money is conserved — no rule creates or destroys any — but the distribution has
gone from a spike at 100 to something with a long right tail. That is the whole
point of the model.

## Setup: agents, networks, globals

`abm_agents()` takes `n` and any number of named columns. A plain value recycles;
a one-sided formula is evaluated once and must return `n` values (or one, to
recycle). Formulas run in order, so a later column can use an earlier one:

```{r}
abm_agents(
  n = 200,
  wtp   = ~rnorm(n, 50, 10),
  offer = ~wtp * 0.8
)
```

A model can have several kinds of agent, declared as a named list. Each keeps its
own columns:

```{r}
market <- abm_setup(agents = list(
  buyers  = abm_agents(n = 5, wtp = ~rnorm(n, 50, 10), offer = ~wtp * 0.8),
  sellers = abm_agents(n = 5, wta = ~rnorm(n, 40, 10), ask   = ~wta * 1.2)
))
market
```

Agent ids are unique across the whole model, and every agent carries a `.group`
column naming its group.

Values shared by the whole population — a bar's last attendance, a bank's ledger
— go in `globals`. They are readable inside every rule:

```{r}
elfarol <- abm_setup(
  agents  = abm_agents(n = 100, threshold = ~runif(n, 40, 80)),
  globals = list(last_attendance = 60)
)
```

A network is an edge list built once and kept for the whole run:

```{r}
abm_network(type = "random", degree = 4)
```

`type = "random"` gives every agent *exactly* `degree` neighbours, using
`igraph::sample_k_regular()`. `degree = 1` therefore pairs the whole population
off permanently.

## Go: a sequence of steps

`abm_go()` takes steps in order and dispatches them by type, not by argument
name. That means a model with several phases per tick is written flat:

```{r}
pd_go <- abm_go(
  # phase 1: play
  abm_match(pair = "random"),
  abm_rules(payoff ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ 3,
    strategy == "defect"    & partner_strategy == "defect"    ~ 1,
    strategy == "defect"    & partner_strategy == "cooperate" ~ 5,
    strategy == "cooperate" & partner_strategy == "defect"    ~ 0
  )),
  # phase 2: imitate whoever did better
  abm_match(pair = "random"),
  abm_rules(strategy ~ if_else(partner_payoff > payoff, partner_strategy, strategy))
)
pd_go
```

The sequence is checked when you build it. Two matches in a row, or a sequence
ending on a match, means a pairing nobody uses, and that is an error rather than a
silent waste:

```{r, error = TRUE}
abm_go(
  abm_match(pair = "random"),
  abm_match(pair = "random"),
  abm_rules(x ~ 1)
)
```

Matching is optional. A model where every agent decides alone needs no
`abm_match()` at all:

```{r}
farol_go <- abm_go(
  abm_rules(go_today ~ last_attendance < threshold),
  abm_global(last_attendance ~ sum(go_today))
)
abm_run(elfarol, farol_go, ticks = 5, seed = 1) |> abm_globals()
```

## What a match gives a rule

For a pair, every rule downstream of the match can see `partner_<col>` for each
of the partner's columns:

```{r}
rumour <- abm_setup(
  agents = abm_agents(n = 100, state = ~c("spreader", rep("ignorant", n - 1)))
)

r <- abm_run(rumour, abm_go(
  abm_match(pair = "random"),
  abm_rules(state ~ case_when(
    state == "ignorant" & partner_state == "spreader" ~ "spreader",
    state == "spreader" & partner_state == "spreader" ~ "stifler",
    state == "spreader" & partner_state == "stifler"  ~ "stifler",
    TRUE ~ state
  ))
), ticks = 100, seed = 1)

table(r$state[r$tick == 100])
```

For a group, rules are evaluated group by group, so an aggregate inside a rule
means "across this agent's group". The public goods game is one line because of
it:

```{r}
pgg <- abm_setup(agents = abm_agents(
  n = 100, contribution = ~sample(c(0, 1), n, replace = TRUE), payoff = 0
))

r <- abm_run(pgg, abm_go(
  abm_match(pair = "random", size = 4),
  abm_rules(payoff ~ sum(contribution) * 2 / 4)
), ticks = 10, seed = 1)

table(r$payoff[r$tick == 10])
```

Grouping is also what makes randomness inside a rule behave: with a match
standing, `sample(x, 1)` is drawn once per pair, not once for everybody.

## Simultaneous or sequential

Every rule in one `abm_rules()` call sees the state at the *start* of the step.
That is the synchronous update agent-based models normally assume, and it is the
one place `abm_rules()` deliberately differs from `dplyr::mutate()`:

```{r}
swap <- abm_setup(agents = abm_agents(n = 3, a = 1, b = 2))
abm_run(swap, abm_go(abm_rules(a ~ b, b ~ a)), ticks = 1)
```

Sometimes that is wrong. If agents compete for a pool that gets *used up*, the
first agent to act has to change what the next one sees. `abm_sequential()`
processes agents one at a time in shuffled order, and its writes to globals are
visible to the agents after it:

```{r}
pot <- abm_setup(agents = abm_agents(n = 20, got = 0), globals = list(pot = 10))

seq_run <- abm_run(pot, abm_go(abm_sequential(
  got ~ if_else(pot > 0, 1, 0),
  pot ~ if_else(pot > 0, pot - 1, pot)
)), ticks = 1, seed = 1)

sum(seq_run$got[seq_run$tick == 1])  # exactly the 10 units that existed
```

With `abm_rules()` instead, all twenty agents would see a full pot and all twenty
would be served. Use `abm_sequential()` only when that difference is the point:
it is slower and harder to reason about.

The order is a fresh shuffle unless you name one. When the order is itself part
of the model — a queue at a counter, a sequential-service constraint — pass
`.order =` and the agents go through in its ascending order instead:

```{r}
queue <- abm_setup(agents = abm_agents(n = 20, place = ~sample(n), got = 0),
                   globals = list(pot = 10))

abm_run(queue, abm_go(abm_sequential(
  got ~ if_else(pot > 0, 1, 0),
  pot ~ if_else(pot > 0, pot - 1, pot),
  .order = place
)), ticks = 1, seed = 1) -> q
sum(q$got[q$tick == 1 & q$place <= 10])   # the front of the queue, every time
```

## Repeating a block

A tick is one pass through `abm_go()`. Some models have a *phase* inside the
tick that has to finish first — an epidemic that burns out before anyone
reconsiders vaccinating, a round of proposals that runs until nobody is
rejected. `abm_repeat()` replays a block until a condition holds:

```{r}
grow <- abm_setup(agents = abm_agents(n = 5, x = 0))
r <- abm_run(grow, abm_go(
  abm_repeat(abm_rules(x ~ x + 1), until = mean(x) >= 4, max = 100)
), ticks = 1, seed = 1)
unique(r$x[r$tick == 1])
```

`until` is checked after each pass, so the block always runs at least once, and
`max` is required — a condition that never becomes true would otherwise hang the
run.

## Changing the population

`abm_birth()` and `abm_death()` are the only steps that add or remove agents.

```{r}
ethno <- abm_setup(agents = abm_agents(
  n = 100,
  strategy = ~sample(c("cooperate", "defect"), n, replace = TRUE),
  resource = 10
))

r <- abm_run(ethno, abm_go(
  abm_match(pair = "random"),
  abm_rules(resource ~ case_when(
    strategy == "cooperate" & partner_strategy == "cooperate" ~ resource + 2,
    strategy == "defect"    & partner_strategy == "cooperate" ~ resource + 4,
    strategy == "cooperate" & partner_strategy == "defect"    ~ resource - 1,
    TRUE ~ resource
  ) - 1),
  abm_birth(when = resource > 20, cost = resource ~ resource / 2),
  abm_death(when = resource <= 0)
), ticks = 20, seed = 1)

table(r$tick)[c(1, 11, 21)]
```

`cost` says what reproduction costs, as ordinary `column ~ expression` formulas
applied to the parent and the newborn alike — halving a resource splits it
between them.

## Reproducibility

Agent-based models are stochastic, so `seed` is an argument to `abm_run()` rather
than something you arrange yourself. It is set locally, so your session's random
state is untouched:

```{r}
set.seed(1); before <- runif(1)
invisible(abm_run(economy, go, ticks = 5, seed = 99))
set.seed(1); after <- runif(1)
identical(before, after)
```

Two runs of the same model with the same seed are identical:

```{r}
a <- abm_run(economy, go, ticks = 20, seed = 42)
b <- abm_run(economy, go, ticks = 20, seed = 42)
identical(as.data.frame(a), as.data.frame(b))
```

There is a sharp edge here worth knowing about. `abm_run()`'s seed fixes the
*run*, not the *model*. If your agents' starting columns are random, they were
drawn when `abm_setup()` was called — before `abm_run()` had a chance to set
anything — so two experiments built the same way can start from different
populations:

```{r}
random_pop <- function() abm_setup(agents = abm_agents(n = 50, x = ~runif(n)))
set.seed(1); one <- abm_run(random_pop(), abm_go(abm_rules(x ~ x)), 1, seed = 5)
set.seed(2); two <- abm_run(random_pop(), abm_go(abm_rules(x ~ x)), 1, seed = 5)
identical(one$x, two$x)
```

Seed both, and the whole experiment reproduces:

```{r}
experiment <- function() {
  m <- abm_setup(agents = abm_agents(n = 50, x = ~runif(n)), seed = 4)
  abm_run(m, abm_go(abm_rules(x ~ x)), ticks = 1, seed = 5)
}
set.seed(1); one <- experiment()
set.seed(2); two <- experiment()
identical(one$x, two$x)
```

One further caveat: a seed fixes a run given a version of the package, not across
versions. Reordering an internal random draw changes the exact numbers without
changing the model, so compare runs by their statistical behaviour rather than by
exact equality when you upgrade.

## Where to go next

`vignette("models")` works through the models the package was designed against,
each with the mechanism it demonstrates.
